Constraints on Cosmic-Ray Particle Populations, Magnetic Fields, and Dark Matter from VHE Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS

### 2011-10-10 ms.tex
Further edits as suggested by G. Tesic

\section{Introduction}
Clusters of galaxies are the largest virialized objects in the Universe, with typical diameters of
a few Mpc and masses on the order of $10^{14}$ to $10^{15} M_{\odot}$. According to the currently
favored hierarchical model of cosmic structure formation, larger objects formed through successive
mergers of smaller objects with galaxy clusters sitting on top of this mass hierarchy
\citep[see][for a review]{article:Voit:2005}. Most of the mass of a cluster is in form of dark
matter (DM), as indicated by galaxy dynamics and gravitational lensing
\citep{article:DiaferioSchindlerDolag:2008}. Baryonic gas making up the intra-cluster medium (ICM)
contributes about 15\% of the total cluster mass and individual galaxies account for the remainder
(about 5\%). The ICM gas mass also comprises a significant fraction of the observable (baryonic)
matter in the Universe.

The ICM is a hot, $T\sim 10^{8}$ K, plasma emitting thermal bremsstrahlung in the soft X-ray regime
\citep[see, e.g.,][]{article:Petrosian:2001}. This plasma has been heated primarily through
collisionless structure-formation shocks that form as a result of the hierarchical merging and
accretion processes. Such shocks and turbulence in the ICM gas in combination with intra-cluster
magnetic fields also provide a means to efficiently accelerate particles \citep[see,
e.g.,][]{article:ColafrancescoBlasi:1998, article:Ryu_etal:2003}. Many clusters do indeed feature
mega-parsec scale halos of non-thermal radio emission, indicative of a population of relativistic
electrons and positrons (leptons) and magnetic fields permeating the ICM
\citep{article:Cassano_etal:2010}. The two theories competing to explain the radio halo are the
``hadronic model'', in which the radio-emitting electrons are produced in inelastic collisions of
cosmic-ray (CR) ions with thermal gas of the ICM \citep{article:Dennison:1980,
  article:EnsslinPfrommerMiniatiSubramanian:2011}, and the ``re-acceleration model'', in which
during states of strong ICM turbulence, e.g. after a cluster merger, interactions of a long-lived
pool of 100-MeV electrons with plasma waves might be efficient enough to accelerate them to energies
($\sim 10$ GeV) sufficient to produce observable radio emission \citep[][and references
therein]{article:BrunettiLazarian:2010}. Recent observations of possibly non-thermal emission from
clusters in the the extreme ultraviolet \citep[EUV; ][]{article:SarazinLieu:1998} and hard X-rays
\citep{article:RephaeliGruber:2002, article:Fusco-Femiano_etal:2004, article:Eckert_etal:2007} may
provide further indications of relativistic particle populations in clusters, although the
interpretation of these observations as non-thermal diffuse emission have been disputed on the basis
of more sensitive observations \citep[see, e.g.,][]{article:Ajello_etal:2009,
  article:Ajello_etal:2010, article:Wik_etal:2009}.

Galaxy clusters have for many years been proposed as sources of gamma rays. If shock acceleration in
the ICM is an efficient process, a population of highly relativistic CR protons and heavy ions is to
be expected in the ICM. The main energy-loss mechanism for CR hadrons at high energies is pion
production through the interaction of CRs with nuclei in the ICM. Pions are short lived and decay;
the charged pions to electrons and positrons,\footnote{A charged pion decays to a muon or antimuon,
  which subsequently decay to an electron or positron as well as neutrinos.} and the neutral pion to
gamma rays. Due to the low density of the ICM ($n_{\mathrm{ICM}}\sim 10^{-3}$ cm$^{-3}$), the large
scales of a cluster and the volume-filling magnetic fields in the ICM, CR hadrons will be confined
in the cluster on timescales comparable to or longer than the Hubble time
\citep[][]{article:Volk_etal:1996, article:Berezinsky_etal:1997} allowing them to accumulate over
time. Already a very modest CR-to-thermal pressure fraction, $X_\CR=\expval{P_{\CR}}/
\expval{P_{\mathrm{th}}}$, of a few per cent implies an observable flux of high-energy gamma rays
\citep[e.g.,][]{article:PfrommerEnsslin:2004b}. For a given CR distribution function, the
hadronically induced gamma-ray flux is directly proportional to $X_{\CR}$.

Hydrostatic estimates of cluster masses, which are determined by balancing the thermal pressure
force and the gravitational force, are biased low by the presence of any substantial non-thermal
pressure component, including a CR pressure contribution. Similarly, a substantial relative CR
pressure can bias the temperature decrement of the cosmic microwave background (CMB) in the
direction of a galaxy cluster due to the Sunyaev-Zel'dovich effect. This could then severely
jeopardize the use of clusters to determine cosmological parameters. Comparing X-ray and optical
potential profiles in the centers of galaxy clusters yields an upper limit of 20-30\% of non-thermal
pressure (that can be composed of CRs, magnetic fields or turbulence) relative to the thermal gas
pressure \citep{article:Churazov_etal:2008, article:Churazov_etal:2010}. An analysis that compares
spatially resolved weak gravitational lensing and hydrostatic X-ray masses for a sample of 18 galaxy
clusters detects a deficit of the hydrostatic mass estimate compared to the lensing mass of $20\%$
at $R_{500}$ -- the radius within which the mean density is 500 times the critical density --
suggesting again a substantial non-thermal pressure contribution on large scales
\citep{article:Mahdavi_etal:2008}. Observing high-energy gamma-ray emission is a complementary
method of constraining the pressure contribution of CRs that is most sensitive to the cluster core
region. However, it assumes that the CR component is fully mixed with the ICM and may not allow to
detect a two-phase structure of CRs and the thermal ICM. A fraction $X_\CR$ of only a few percent is
required in order to produce a gamma-ray flux observable with current generation gamma-ray
telescopes rendering this technique at least as sensitive as the dynamical and hydrostatic method
(which are more general in that they are sensitive to any non-thermal pressure component).

@@

A third mechanism for gamma-ray production in a galaxy cluster is self-annihilation of a DM
particle, e.g. a weakly interacting massive particle (WIMP). As already mentioned, about 80\% of
the cluster mass is in the form of dark matter. The gamma-ray emissivity from DM annihilations is
proportional to the square of the DM density integrated along the line of sight, making galaxy clusters
potentially good candidates for DM searches \citep{article:EvansFerrerSarkar:2004,
article:BergstromHooper:2006, article:PinzkePfrommerBergstrom2009, article:Cuesta_etal:2011},
despite their large distances compared to other common targets for DM searches, such as dwarf
spheroidal galaxies \citep{article:Strigari_etal:2007, article:Acciari_etal:2010,article:Aliu_etal:2009}
or the Galactic center \citep{article:Kosack_etal:2004, article:Aharonian_etal:2006,
article:Aharonian_etal:2009b}.

While several observations of clusters of galaxies have been made with satellite-borne and
ground-based gamma-ray telescopes, a detection of gamma-ray emission from a cluster has yet to be
made. Observations with EGRET \citep{article:Sreekumar_etal:1996, article:Reimer_etal:2003} and the
Fermi Gamma-ray Space Telescope \citep{article:Ackermann_etal:2010} have provided upper limits on the
gamma-ray fluxes (typically $\sim10^{-9}$ ph cm$^{2}$ s$^{-1}$ for Fermi observations) for several
galaxy clusters in the MeV to GeV band. Upper limits on the very-high-energy (VHE) gamma-ray flux
from a few selected clusters, including the Coma cluster, have been provided by observations with
ground-based imaging atmospheric Cherenkov telescopes \citep[IACTs;][]{article:Perkins_etal:2006,
inproc:Perkins_etal:2008, article:Aharonian_etal:2009a, article:Aleksic_etal:2010}.

The Coma cluster of galaxies (ACO 1656) is one of the most thoroughly studied clusters over all
wavelengths \citep{article:Voges_etal:1999}. Located at a distance of about 100 Mpc
\citep[$z=0.023$;][]{article:StrubleRood:1999}, it is one of the closest massive clusters \citep[$M
\sim10^{15}M_{\odot}$;][]{article:Smith:1983, article:Kubo_etal:2008}. It hosts both a giant radio
halo \citep{article:Giovannini_etal:1993,article:Thierbach_etal:2003} and peripheral radio relic,
which appears connected to the radio halo with a ``diffuse'' bridge \citep[see discussion
in][]{article:BrownRudnick:2010}. It has been suggested \citep{article:Ensslin_etal:1998} and
successively demonstrated by cosmological simulations, which model the non-thermal emission
processes, \citep{article:PfrommerEnsslinSpringel:2008, article:Pfrommer:2008,
  article:Battaglia_etal:2009, article:Skillman_etal:2011}, that the relic could well be an infall
shock. Extended soft thermal X-ray (SXR) emission is evident from the ROSAT all-sky survey in the
0.1 to 2.4 keV band \citep{article:BrielHenryBohringer:1992}. Observations with XMM-Newton
\citep{article:Briel_etal:2001} revealed substructure in the X-ray halo supported by substantial
turbulent pressure of at least $\sim 10 \%$ of the total pressure
\citep{article:Schuecker_etal:2004}. The Coma cluster is a natural candidate for gamma-ray
observations.

@@

\section{VERITAS Observations and Analysis}
The VERITAS gamma-ray detector \citep{article:Weekes_etal:2002} is an array of four 12 m diameter
imaging atmospheric Cherenkov telescopes \citep[IACTs;][]{article:Holder_etal:2006} located at an
altitude of 1268 m a.s.l. at the Fred Lawrence Whipple Observatory in southern Arizona
(31$^{\circ}$~40\arcmin~30\arcsec~N, 110$^{\circ}$~57\arcmin~07\arcsec~W). Each of the telescopes
is equipped with a 499 pixel camera covering a 3.5$^{\circ}$ field of view. The array, completed in
the spring of 2007, is designed to detect gamma-ray emission from astrophysical objects in the
energy range from 100 GeV to more than 30 TeV. The energy resolution is about 15\% and the angular
resolution (68\% containment) is about 0.1$^{\circ}$ per event at 1 TeV. The sensitivity of the
array allows for detection of a point source with a flux of 1\% of the steady Crab Nebula flux
above 300 GeV at the $5\sigma$ level in under 30 hours.\footnote{The integral flux sensitivity
above 300 GeV was improved by about 30\% with the move of one telescope in the summer of 2009. The
move had no impact on this data set.}

@@

Gamma-ray like events are separated from the CR background by imposing cuts on scaled parameters
\citep{article:Aharonian_etal:1997, article:Krawczynski_etal:2006} calculated from a parametrized
moment analysis of the shower images \citep{inproc:Hillas:1985}. The gamma-hadron separation cuts
used in this analysis were optimized a priori for a weak point source (3\% Crab Nebula flux level)
and a photon index of 2.4 using data taken on the Crab Nebula during the same epoch. Because the VHE
gamma-ray spectrum for the Coma cluster is expected to be hard, photon index of about 2.3
\citep{article:PinzkePfrommer:2010}, these cuts are suitable for the analysis of the Coma cluster
data set.

@@

\subsection{VERITAS Results}
No significant excess of VHE gamma-rays from the Coma cluster was detected with VERITAS, as
illustrated by the $\theta^{2}$ distribution shown in Figure \ref{fig:thetasq}. The $\theta^{2}$
distribution is a radial comparison of the ON-source event distribution with the estimated
background. It extends out to a radius of $0.42^{\circ}$ to cover all regions of interest
considered in this work. A 99\% confidence level upper limit is calculated for each region of
interest using the method described by \citet{article:Rolke_etal:2005} assuming a Gaussian
background. A lower bound of zero is imposed on the signal rate to handle the negative excesses
seen for several of the regions of interest. Figure \ref{fig:sigdist} shows that the VERITAS data
is well fit by a Gaussian with a mean close to zero and a standard deviation within a few percent
of unity.

@@

A few remarks are in order. (1) For the VERITAS limits, the hardest
CR spectral indices correspond to the tightest limits on $B_{0}$, because the CR flux is constrained
around 1 TeV and a comparably small fraction of CRs at 100 GeV would be available to produce
radio-emitting electrons. A high magnetic field would be required to match the observed
synchrotron emission. The opposite is true for the \emph{Fermi} upper limits at 1 GeV, which
probe CRs around a pivot point of 8 GeV: a soft CR spectral index implies a comparably small
fraction of CRs at 100 GeV and hence a strong magnetic field is needed to match the observed
synchrotron flux. (2) For a steeper
magnetic decline (larger $\alpha_{B}$), the CR number density needs to be larger to match the
observed radio emission profiles which would yield a higher gamma-ray flux so that the upper limits
are more constraining. This implies tighter lower limits for $B_{0}$.  (3) Interestingly, in all
cases, the 0.4$^{\circ}$-aperture limits that are the most constraining. For a given magnetic
realization a substantially increasing relative CR pressure profile is needed to match the
observed radio profiles, and therefore that CR realization produces a larger flux within 0.4$^{\circ}$
in comparison with the simplified CR model ($X_{\CR} = \rmn{const.}$), for which the
0.2$^{\circ}$-aperture limits are more constraining in the case of VERITAS. Physically, the large
CR pressure in the cluster periphery may arise from CR streaming into the large available phase
space in the outer regions.

@@

Given the upper limit on the observed gamma-ray rate,
$R_{\gamma}(99\%\ \mathrm{CL}) = N_{\gamma}(99\%\ \mathrm{CL}) / T_{\mathrm{obs}}$, we can place
constraints on the WIMP parameter space $(m_{\chi}, \expval{\sigma v})$. From Eq.
(\ref{eqn:WIMPflux}) we find
\begin{equation}
\expval{\sigma v}(99\%\ \mathrm{CL}) <
R_{\gamma}(99\%\ \mathrm{CL})\, \frac{8\pi m_{\chi}^{2}}{J(\Delta\Omega)}\,
\frac{1}{\int^{m_{\chi}} dE\ A_{\mathrm{eff}}\,N_\gamma (E)}
\end{equation}
Because the self-annihilation of a WIMP is a two-body process, the astrophysical factor
$J(\Delta\Omega)$ can be expressed as the line-of-sight integral of the DM density squared
\begin{equation}
J(\Delta\Omega)=\left(\frac{1}{\rho_{c}^{2}R_{H}}\right)
\int_{\Delta\Omega}d\Omega\int d\lambda\ \rho^{2}(\lambda,\Omega),
\end{equation}
where $\lambda$ represents the line-of-sight. Here,
$J(\Delta\Omega)$ has been normalized to the square of the critical density, $\rho_{c}$, and the
Hubble radius, $R_{H}$. In this work we have adopted the DM density profile by
\citet{article:NavarroFrenkWhite:1997},
\begin{equation}
\rho_{\chi}(r)=\rho_{s}\left(\frac{r}{r_{s}}\right)^{-1}\left(1+\frac{r}{r_{s}}\right)^{-2},
\end{equation}
which parametrizes the distribution of DM by the scale radius $r_{s}$ and the scale density
$\rho_{s}$. Using weak-lensing measurements of the virial mass in the Coma cluster
\citep{article:Gavazzi_etal:2009} we find for these parameters $r_{s}=0.654$ Mpc and
$\rho_{s}=4.4\times 10^{4}$ M$_{\odot}$/Mpc$^{3}$. Table \ref{table:astrofactor} lists the
astrophysical factors calculated for the different VERITAS apertures considered in this work.
It also lists the astrophysical factor calculated for the background region, which is used
to estimate the gamma-ray contamination due to DM in the background region.

@@

We employed various approaches to constrain the CR population and magnetic field distribution that
are complementary in their assumptions and hence well suited in assessing underlying Bayesian
priors in the models. (1) We use a simplified ``isobaric CR model'' that is characterized by a
constant CR-to-thermal pressure fraction and has a power-law momentum spectrum.  While this model
is physically not justified a priory, it is simple and widely used in the literature and captures
some aspects of more elaborate models such as (2) the simulation-based analytical approach of
\citet{article:PinzkePfrommer:2010}. This is a ``first principle approach'' that predicts the CR
distribution spectrally and spatially for a given set of assumptions (namely the physics that was
simulated). It is powerful since it only requires the density profile as input due to the
approximate universality of the CR distribution (when neglecting CR diffusion and streaming which
appears, however, necessary to explain the radio halo bimodality). (3) We finally use a pragmatic
approach which models the CR and magnetic distribution such as to reproduce the observed emission
profile of the Coma radio halo. While this approach is also physically not justified, it is
powerful as it shows what the ``correct'' model has to achieve and can point into the direction of
the relevant physics.

>>>>

\section{Introduction}
Clusters of galaxies are the largest virialized objects in the Universe, with typical diameters of
a few Mpc and masses on the order of $10^{14}$ to $10^{15} M_{\odot}$. According to the currently
favored hierarchical model of cosmic structure formation, larger objects formed through successive
mergers of smaller objects with galaxy clusters sitting on top of this mass hierarchy
\citep[see][for a review]{article:Voit:2005}. Most of the mass ($\sim$80\%) in a cluster is in form
of dark matter (DM), as indicated by galaxy dynamics and gravitational lensing
\citep{article:DiaferioSchindlerDolag:2008}. Baryonic gas making up the intra-cluster medium (ICM)
contributes about 15\% of the total cluster mass and individual galaxies account for the remainder
(about 5\%). The ICM gas mass also comprises a significant fraction of the observable (baryonic)
matter in the Universe.

The ICM is a hot, $T\sim 10^{8}$ K, plasma emitting thermal bremsstrahlung in the soft X-ray regime
\citep[see, e.g.,][]{article:Petrosian:2001}. This plasma has been heated primarily through
collisionless structure-formation shocks that form as a result of the hierarchical merging and
accretion processes. Such shocks and turbulence in the ICM gas in combination with intra-cluster
magnetic fields also provide a means to efficiently accelerate particles \citep[see,
e.g.,][]{article:ColafrancescoBlasi:1998, article:Ryu_etal:2003}. Many clusters feature
mega-parsec scale halos of non-thermal radio emission, indicative of a population of relativistic
electrons and positrons (leptons) and magnetic fields permeating the ICM
\citep{article:Cassano_etal:2010}. The two theories competing to explain the radio halo are the
``hadronic model'', in which the radio-emitting electrons are produced in inelastic collisions of
cosmic-ray (CR) ions with thermal gas of the ICM \citep{article:Dennison:1980,
  article:EnsslinPfrommerMiniatiSubramanian:2011}, and the ``re-acceleration model'', in which
during states of strong ICM turbulence, e.g. after a cluster merger, interactions of a long-lived
pool of 100-MeV electrons with plasma waves might be efficient enough to accelerate them to energies
($\sim 10$ GeV) sufficient to produce observable radio emission \citep[][and references
therein]{article:BrunettiLazarian:2010}. Recent observations of possibly non-thermal emission from
clusters in the the extreme ultraviolet \citep[EUV; ][]{article:SarazinLieu:1998} and hard X-rays
\citep{article:RephaeliGruber:2002, article:Fusco-Femiano_etal:2004, article:Eckert_etal:2007} may
provide further indications of relativistic particle populations in clusters, although the
interpretation of these observations as non-thermal diffuse emission have been disputed on the basis
of more sensitive observations \citep[see, e.g.,][]{article:Ajello_etal:2009,
  article:Ajello_etal:2010, article:Wik_etal:2009}.

Galaxy clusters have, for many years, been proposed as sources of gamma rays. If shock acceleration
in the ICM is an efficient process, a population of highly relativistic CR protons and heavy ions
is to be expected in the ICM. The main energy-loss mechanism for CR hadrons at high energies is pion
production through the interaction of CRs with nuclei in the ICM. Pions are short lived and decay;
the charged pions to electrons and positrons,\footnote{A charged pion decays to a muon or antimuon,
  which subsequently decay to an electron or positron as well as neutrinos.} and the neutral pion to
gamma rays. Due to the low density of the ICM ($n_{\mathrm{ICM}}\sim 10^{-3}$ cm$^{-3}$), the large
scales of a cluster and the volume-filling magnetic fields in the ICM, CR hadrons will be confined
in the cluster on timescales comparable to or longer than the Hubble time
\citep[][]{article:Volk_etal:1996, article:Berezinsky_etal:1997} allowing them to accumulate over
time. For a given CR distribution function, the hadronically induced gamma-ray flux is directly
proportional to the CR-to-thermal pressure fraction, $X_\CR=\expval{P_{\CR}}/
\expval{P_{\mathrm{th}}}$. Already a very modest $X_{\CR}$ of a few percent implies an observable
flux of high-energy gamma rays \citep[e.g.,][]{article:PfrommerEnsslin:2004b}.

Hydrostatic estimates of cluster masses, which are determined by balancing the thermal pressure
force and the gravitational force, are biased low by the presence of any substantial non-thermal
pressure component, including a CR pressure contribution. Similarly, a substantial relative CR
pressure can bias the temperature decrement of the cosmic microwave background (CMB) in the
direction of a galaxy cluster due to the Sunyaev-Zel'dovich effect. This could then severely
jeopardize the use of clusters to determine cosmological parameters. Comparing X-ray and optical
potential profiles in the centers of galaxy clusters yields an upper limit of 20-30\% of non-thermal
pressure (that can be composed of CRs, magnetic fields or turbulence) relative to the thermal gas
pressure \citep{article:Churazov_etal:2008, article:Churazov_etal:2010}. An analysis that compares
spatially resolved weak gravitational lensing and hydrostatic X-ray masses for a sample of 18 galaxy
clusters detects a deficit of the hydrostatic mass estimate compared to the lensing mass of $20\%$
at $R_{500}$ -- the radius within which the mean density is 500 times the critical density of the
Universe -- suggesting again a substantial non-thermal pressure contribution on large scales
\citep{article:Mahdavi_etal:2008}. Observing high-energy gamma-ray emission is a complementary
method of constraining the pressure contribution of CRs that is most sensitive to the cluster core
region. However, it assumes that the CR component is fully mixed with the ICM and may not allow for
a detection of a two-phase structure of CRs and the thermal ICM. A fraction $X_\CR$ of only a few 
percent is required in order to produce a gamma-ray flux observable with current generation of
gamma-ray telescopes rendering this technique at least as sensitive as the dynamical and
hydrostatic method (which are more general in that they are sensitive to any non-thermal pressure
component).

@@

A third mechanism for gamma-ray production in a galaxy cluster is self-annihilation of a DM
particle, e.g. a weakly interacting massive particle (WIMP). As already mentioned, about 80\% of
the cluster mass is in the form of dark matter. The gamma-ray emissivity from DM annihilations is
proportional to the square of the DM density integrated along the line of sight, making galaxy
clusters potentially good candidates for DM searches \citep{article:EvansFerrerSarkar:2004,
article:BergstromHooper:2006, article:PinzkePfrommerBergstrom2009, article:Cuesta_etal:2011},
despite their large distances compared to other common targets for DM searches, such as dwarf
spheroidal galaxies \citep{article:Strigari_etal:2007, article:Acciari_etal:2010,
article:Aliu_etal:2009} or the Galactic center \citep{article:Kosack_etal:2004,
article:Aharonian_etal:2006, article:Aharonian_etal:2009b, article:Abramowski_etal:2011}.

While several observations of clusters of galaxies have been made with satellite-borne and
ground-based gamma-ray telescopes, a detection of gamma-ray emission from a cluster has yet to be
made. Observations with EGRET \citep{article:Sreekumar_etal:1996, article:Reimer_etal:2003} and the
Fermi Gamma-ray Space Telescope \citep{article:Ackermann_etal:2010} have provided upper limits on
the gamma-ray fluxes (typically $\sim10^{-9}$ ph cm$^{2}$ s$^{-1}$ for Fermi observations) for
several galaxy clusters in the MeV to GeV band. Upper limits on the very-high-energy (VHE)
gamma-ray flux from a select few clusters, including the Coma cluster, have been provided by
observations with ground-based imaging atmospheric Cherenkov telescopes
\citep[IACTs;][]{article:Perkins_etal:2006, inproc:Perkins_etal:2008, article:Aharonian_etal:2009a,
article:Aleksic_etal:2010}.

The Coma cluster of galaxies (ACO 1656) is one of the most thoroughly studied clusters over all
wavelengths \citep{article:Voges_etal:1999}. Located at a distance of about 100 Mpc
\citep[$z=0.023$;][]{article:StrubleRood:1999}, it is one of the closest massive clusters \citep[$M
\sim10^{15}M_{\odot}$;][]{article:Smith:1983, article:Kubo_etal:2008}. It hosts both a giant radio
halo \citep{article:Giovannini_etal:1993,article:Thierbach_etal:2003} and peripheral radio relic,
which appears connected to the radio halo with a ``diffuse'' bridge \citep[see discussion
in][]{article:BrownRudnick:2010}. It has been suggested \citep{article:Ensslin_etal:1998} and
successively demonstrated by cosmological simulations, which model the non-thermal emission
processes \citep{article:PfrommerEnsslinSpringel:2008, article:Pfrommer:2008,
article:Battaglia_etal:2009, article:Skillman_etal:2011}, that the relic could well be an infall
shock. Extended soft thermal X-ray (SXR) emission is evident from the ROSAT all-sky survey in the
0.1 to 2.4 keV band \citep{article:BrielHenryBohringer:1992}. Observations with XMM-Newton
\citep{article:Briel_etal:2001} revealed substructure in the X-ray halo supported by substantial
turbulent pressure of at least $\sim 10 \%$ of the total pressure
\citep{article:Schuecker_etal:2004}. The Coma cluster is a natural candidate for gamma-ray
observations.

@@

\section{VERITAS Observations and Analysis}
The VERITAS gamma-ray detector \citep{article:Weekes_etal:2002} is an array of four 12 m diameter
imaging atmospheric Cherenkov telescopes \citep[IACTs;][]{article:Holder_etal:2006} located at an
altitude of $\sim$1250 m a.s.l. at the Fred Lawrence Whipple Observatory in southern Arizona
(31$^{\circ}$~40\arcmin~30\arcsec~N, 110$^{\circ}$~57\arcmin~07\arcsec~W). Each of the telescopes
is equipped with a 499 pixel camera covering a 3.5$^{\circ}$ field of view. The array, completed in
the spring of 2007, is designed to detect gamma-ray emission from astrophysical objects in the
energy range from 100 GeV to more than 30 TeV. The energy resolution is about 15\% and the angular
resolution (68\% containment) is about 0.1$^{\circ}$ per event at 1 TeV. The sensitivity of the
array allows for detection of a point source with a flux of 1\% of the steady Crab Nebula flux
above 300 GeV at the $5\sigma$ level in under 30 hours.\footnote{The integral flux sensitivity
above 300 GeV was improved by about 30\% with the move of one telescope in the summer of 2009. The
move had no impact on this data set.}

@@

Gamma-ray like events are separated from the CR background by imposing cuts on scaled parameters
\citep{article:Aharonian_etal:1997, article:Krawczynski_etal:2006} calculated from a parametrized
moment analysis of the shower images \citep{inproc:Hillas:1985}. The gamma-hadron separation cuts
used in this analysis were optimized a priori for a weak point source (3\% Crab Nebula flux level)
and a photon index of 2.4 using data taken on the Crab Nebula during the same epoch. Because the VHE
gamma-ray spectrum for the Coma cluster is expected to be hard, with a photon index of about 2.3
\citep{article:PinzkePfrommer:2010}, these cuts are suitable for the analysis of the Coma cluster
data set.

@@

\subsection{VERITAS Results}
No significant excess of VHE gamma-rays from the Coma cluster was detected with VERITAS, as
illustrated by the $\theta^{2}$ distribution shown in Figure \ref{fig:thetasq}. The $\theta^{2}$
distribution is a radial comparison of the ON-source event distribution with the estimated
background. It extends out to a radius of 0.42 square degrees to cover all regions of interest
considered in this work. A 99\% confidence level upper limit is calculated for each region of
interest using the method described by \citet{article:Rolke_etal:2005} assuming a Gaussian
background. A lower bound of zero is imposed on the signal rate to handle the negative excesses
seen for several of the regions of interest. Figure \ref{fig:sigdist} shows that the VERITAS data
is well fit by a Gaussian with a mean close to zero and a standard deviation within a few percent
of unity.

@@

A few remarks are in order. (1) For the VERITAS limits, the hardest
CR spectral indices correspond to the tightest limits on $B_{0}$, because the CR flux is constrained
around 1 TeV and a comparably small fraction of CRs at 100 GeV would be available to produce
radio-emitting electrons. A high magnetic field would be required to match the observed
synchrotron emission. The opposite is true for the \emph{Fermi} upper limits at 1 GeV, which
probe CRs around a pivot point of 8 GeV: a soft CR spectral index implies a comparably small
fraction of CRs at 100 GeV and hence a strong magnetic field is needed to match the observed
synchrotron flux. (2) For a steeper
magnetic decline (larger $\alpha_{B}$), the CR number density needs to be larger to match the
observed radio emission profiles which would yield a higher gamma-ray flux so that the upper limits
are more constraining. This implies tighter lower limits for $B_{0}$. (3) Interestingly, in all
cases, the 0.4$^{\circ}$-aperture limits are the most constraining. For a given magnetic
realization a substantially increasing relative CR pressure profile is needed to match the
observed radio profiles, and therefore that CR realization produces a larger flux within 
0.4$^{\circ}$ in comparison with the simplified CR model ($X_{\CR} = \rmn{const.}$), for which the
0.2$^{\circ}$-aperture limits are more constraining in the case of VERITAS. Physically, the large
CR pressure in the cluster periphery may arise from CR streaming into the large available phase
space in the outer regions.

@@

Given the upper limit on the observed gamma-ray rate,
$R_{\gamma}(99\%\ \mathrm{CL}) = N_{\gamma}(99\%\ \mathrm{CL}) / T_{\mathrm{obs}}$, we can place
constraints on the WIMP parameter space $(m_{\chi}, \expval{\sigma v})$. From Eq.
(\ref{eqn:WIMPflux}) we find
\begin{equation}
\expval{\sigma v}(99\%\ \mathrm{CL}) <
R_{\gamma}(99\%\ \mathrm{CL})\, \frac{8\pi m_{\chi}^{2}}{J(\Delta\Omega)}\,
\frac{1}{\int^{m_{\chi}} dE\ A_{\mathrm{eff}}\,N_\gamma (E)},
\end{equation}
where $A_{\mathrm{eff}}$ is the effective area of the gamma-ray detector.
Because the self-annihilation of a WIMP is a two-body process, the astrophysical factor
$J(\Delta\Omega)$ can be expressed as the line-of-sight integral of the DM density squared
\begin{equation}
J(\Delta\Omega)=\left(\frac{1}{\rho_{c}^{2}R_{H}}\right)
\int_{\Delta\Omega}d\Omega\int d\lambda\ \rho^{2}(\lambda,\Omega),
\end{equation}
where $\lambda$ represents the line-of-sight. Here,
$J(\Delta\Omega)$ has been normalized to the square of the critical density, $\rho_{c}$, and the
Hubble radius, $R_{H}$. In this work we have adopted the DM density profile by
\citet{article:NavarroFrenkWhite:1997},
\begin{equation}
\rho_{\chi}(r)=\rho_{s}\left(\frac{r}{r_{s}}\right)^{-1}\left(1+\frac{r}{r_{s}}\right)^{-2},
\end{equation}
which parametrizes the distribution of DM by the scale radius $r_{s}$ and the scale density
$\rho_{s}$. Using weak-lensing measurements of the virial mass in the Coma cluster
\citep{article:Gavazzi_etal:2009} we find for these parameters $r_{s}=0.654$ Mpc and
$\rho_{s}=4.4\times 10^{4}$ M$_{\odot}$/Mpc$^{3}$. Table \ref{table:astrofactor} lists the
astrophysical factors calculated for the different VERITAS apertures considered in this work.
It also lists the astrophysical factor calculated for the background region, which is used
to estimate the gamma-ray contamination due to DM in the background region.

@@

We employed various approaches to constrain the CR population and magnetic field distribution that
are complementary in their assumptions and hence well suited in assessing underlying Bayesian
priors in the models. (1) We use a simplified ``isobaric CR model'' that is characterized by a
constant CR-to-thermal pressure fraction and has a power-law momentum spectrum.  While this model
is physically not justified a priory, it is simple and widely used in the literature and captures
some aspects of more elaborate models such as (2) the simulation-based analytical approach of
\citet{article:PinzkePfrommer:2010}. This is a ``first principle approach'' that predicts the CR
distribution spectrally and spatially for a given set of assumptions (namely the physics that was
simulated). It is powerful since it only requires the density profile as input due to the
approximate universality of the CR distribution (when neglecting CR diffusion and streaming which
appears necessary to explain the radio halo bimodality). (3) We finally use a pragmatic
approach which models the CR and magnetic distribution in order to reproduce the observed emission
profile of the Coma radio halo. While this approach is also physically not justified, it is
powerful as it shows what the ``correct'' model has to achieve and can point into the direction of
the relevant physics.
